Genetic draft and the evolution of ``irreducible complexity"

### 2018-03-19 ms/valley_crossing.tex

There are essentially three components to this process.
First, a single mutant lineage must appear, which occurs with rate $N\mu$.
During the lifetime of this mutant lineage, the complex adaptation must appear and establish, that is, rise to a high enough frequency that fixation is almost guaranteed, which occurs at rate $\mu p_{\mathrm fix}$.
Finally, the complex adaptation must sweep to fixation.
The first two steps can be folded in together and considered as one process, so that what is relevant is the expected time until a single mutant lineage arises that is destined to produce a successful double mutant.
In this way, $\mathbb{E}\left[ T \right] = \mathbb{E} \left[ T_0 + T_1 \right]$, where $T_0$ is the time to the first successful bubble and $T_1$ is the sweep time.
We assume that $T_0 \gg T_1$, so that the wait time is dominated by the first term.

We refer to single mutant lineages as ``bubbles".
Let $n(t)$ be the total number of single mutant individuals in the bubble, and suppose it appears at time $t_0$ and goes extinct at time $T$.
Then $W = \int_{t_0}^T n(t)dt$ is the ``weight" of the bubble, representing the total number of chances there are for a successful double mutant to appear.
Most of the time, a bubble will simply arise and go extinct.
However, with probability $1-e^{-\mu p_{\mathrm fix} W}$, the bubble will give rise to a successful double mutant.
Computing the expected value $\phi = \left< 1-e^{\mu p_{\mathrm fix} W} \right>$ thus gives the probability that a particular bubble will lead to a successful valley crossing.

One difficulty in computing the rate of tunneling in the high $N\sigma$ case is that the expected bubble size and fixation probability will both turn out to depend on the background fitness of an individual.
That is to say, one cannot simply compute $p_{\mathrm{fix}}$ and the weight distribution by themselves but rather must convolute both of them over the background fitness.
Individuals near the nose of the distribution will tend to give rise to longer-lived lineages, and because they are near the nose, the complex adaptation will therefore likewise be more likely to fix.
As a result, $W$ will depend on $x$ and $\delta$, and $p_{\mathrm fix}$ will depend on $x$ and $s$.
These will have to be integrated over the fitness distribution $f(x)$ in order to obtain the total crossing probability $\Phi(s,\delta) = \int_{-\infty}^\infty \phi(x,s,\delta) f(x) dx$.
The expected valley crossing time then becomes $1/N\mu\Phi$.

We begin by considering the dynamics of $p_{\mathrm fix}$ before moving on to consider the distribution of bubble sizes.
Let $p = p_{\mathrm fix}(x,s,t)$ be the fixation probability of the double mutant as a function of the absolute fitness $x$.
In the absence of any perturbations, the evolution of $p$ can be described by a birth-death process, yielding
\begin{equation}
-\partial_t p(x,s,t) = Bp(x,s,t) - (B+D)p(x,s,t)^2
\end{equation}
\citep{barton_1995, good_2012}, with $B$ and $D$ the birth and death rates respectively.
The only dependence of $p$ on $t$ is via the time-changing mean fitness $vt$.
Defining $X = x - vt$, we then have $\partial_t p(x-vt) = -v\partial_X p(X)$.
This can be rewritten as a wave equation
\begin{equation}
v \partial_X p(X,s) = (X+s)p(X,s) - (1+X+s)p(X,s)^2.
\end{equation}
This is a Bernoulli equation and is readily solved, but the only appropriate boundary condition yields $p = 0$ everywhere.
This is intuitive: the inexorable advance of the mean fitness implies that all lineages are ultimately doomed to extinction \emph{unless} they can increase their own fitness either via recombination or mutation.
We thus need to amend the model to include a mutation term.
Let $U$ be the mutation rate of the genetic background and $\epsilon$ be the fitness effect of a mutation, distributed as $\rho(\epsilon)$.
Then we have
\begin{equation}
v \partial_X p(X,s) = (X+s)p(X,s) - (1+X+s)p(X,s)^2 + U\int_0^\infty \rho(\epsilon) \left[ p(X+\epsilon,s)-p(X,s)\right] d\epsilon,
\end{equation}
where $\epsilon$ is now the fitness effect of a mutation in the background distribution, $\rho(\epsilon)$ is the distribution of mutational effects, and $U$ is the mutation rate of the background.
Assuming that $s$ is small compared to $\sigma$, a matched asymptotics analysis yields
\begin{equation}
p(X,s) =
\begin{cases}
0 &\mathrm{~if~} X+s < 0, \\
X_c e^{\frac{(X+s)^2-X_c^2}{2\sigma^2}} &\mathrm{~if~} 0 < X+s < X_c, \\
X+s &\mathrm{~if~} X+s > X_c.
\end{cases}
\end{equation}
(If $s$ is large compared to $\sigma$, then the standard Haldane result, $p(s) \sim s$, obtains.)
Effectively, this means the fixation probability is \emph{zero} if the double mutant lineage's fitness is less than the average for the population: if the fitness is less than some critical value $X_c$, then the lineage must generate a background mutant in order to fix, and if it is above $X_c$, the fixation probability is determined by the lineage's ability to survive drift and establish.
The value of $X_c$ is determined by the shape of $\rho(\epsilon)$.
Suppose $\rho$ is distributed exponentially: $\rho(\epsilon) = \frac{1}{\bar{\epsilon}} e^{-\epsilon/\bar{\epsilon}}$.
Then $X_c$ is determined by solving
\begin{equation}
2 = \frac{\sqrt{2\pi}\sigma U}{\bar{\epsilon}} \left[ \frac{1}{X_c - \frac{\sigma^2}{\bar{\epsilon}}} + \frac{\bar{\epsilon}}{\sigma^2} \left( 1+ \frac{\bar{\epsilon}}{X_c} \right) \right] e^{\frac{(X_c - \sigma^2/\bar{\epsilon})^2}{2\sigma^2}}. 
\end{equation}

We now consider the behavior of single mutant bubbles.
Since the tunneling probability for a lineage with background fitness $x$ is given by $\phi = \left< 1 - e^{-\mu p_{\mathrm fix} W} \right>$, we do not need to compute the full distribution of $W$ but rather only its Laplace transform: $\mathcal{L}\left[ p(w) \right] = \int_0^\infty e^{-zw} p(w) dw$, with $z = \mu p_{\mathrm fix}$.
We proceed to derive an equation for the time evolution of the Laplace transform.

@@

To test our model, we perform forward simulations using a customized version of \texttt{FFPopSim} \citep{zanini_2012}, a discrete time forward simulation package.
The simulation code is available from the authors upon request or via \texttt{GitHub}.
We initialize a wild type population of $N$ individuals consisting of haploid genomes of length $L = 200$.
For the genetic background, we make use of an \emph{ersatz} ``infinite sites" model, where any time a locus becomes monomorphic, a mutation is injected into a random individual in the population.
Background fitness effects are drawn from an exponential distribution, and the background fitness variance is set to a constant value $\sigma^2$ by manually adjusting the selection coefficients every generation.
In this way, $U$ and $\bar{\epsilon}$ are not parameters of our simulation model but must be directly measured from simulations: they are constrained by $L$ and $\sigma$.

@@

@@

>>>>

There are essentially three components to this process.
First, a single mutant lineage must appear, which occurs with rate $N\mu$.
During the lifetime of this mutant lineage, the complex adaptation must appear and establish, that is, rise to a high enough frequency that fixation is almost guaranteed, which occurs at rate $\mu p_{\mathrm {fix}}$.
Finally, the complex adaptation must sweep to fixation.
The first two steps can be folded in together and considered as one process, so that what is relevant is the expected time until a single mutant lineage arises that is destined to produce a successful double mutant.
In this way, $\mathbb{E}\left[ T \right] = \mathbb{E} \left[ T_0 + T_1 \right]$, where $T_0$ is the time to the first successful bubble and $T_1$ is the sweep time.
We assume that $T_0 \gg T_1$, so that the wait time is dominated by the first term.

We refer to single mutant lineages as ``bubbles".
Let $n(t)$ be the total number of single mutant individuals in the bubble, and suppose it appears at time $t_0$ and goes extinct at time $T$.
Then $W = \int_{t_0}^T n(t)dt$ is the ``weight" of the bubble, representing the total number of chances there are for a successful double mutant to appear.
Most of the time, a bubble will simply arise and go extinct.
However, with probability $1-e^{-\mu p_{\mathrm{fix}} W}$, the bubble will give rise to a successful double mutant.
Computing the expected value $\phi = \left< 1-e^{\mu p_{\mathrm {fix}} W} \right>$ thus gives the probability that a particular bubble will lead to a successful valley crossing.
One difficulty in computing the rate of tunneling in the high $N\sigma$ case is that the expected bubble size and fixation probability will both turn out to depend on the background fitness of an individual.
That is to say, one cannot simply compute $p_{\mathrm{fix}}$ and the weight distribution by themselves but rather must convolute both of them over the background fitness.
Individuals near the nose of the distribution will tend to give rise to longer-lived lineages, and because they are near the nose, the complex adaptation will therefore likewise be more likely to fix.
As a result, $W$ will depend on $x$ and $\delta$, and $p_{\mathrm fix}$ will depend on $x$ and $s$.
These will have to be integrated over the fitness distribution $f(x)$ in order to obtain the total crossing probability $\Phi(s,\delta) = \int_{-\infty}^\infty \phi(x,s,\delta) f(x) dx$.
The expected valley crossing time then becomes $1/N\mu\Phi$.

We begin by considering the dynamics of $p_{\mathrm {fix}}$ before moving on to consider the distribution of bubble sizes.
Let $p = p_{\mathrm fix}(x,s,t)$ be the fixation probability of the double mutant as a function of the absolute fitness $x$.
In the absence of any perturbations, the evolution of $p$ can be described by a birth-death process, yielding
\begin{equation}
-\partial_t p(x,s,t) = Bp(x,s,t) - (B+D)p(x,s,t)^2
\end{equation}
\citep{barton_1995, good_2012}, with $B$ and $D$ the birth and death rates respectively.
The only dependence of $p$ on $t$ is via the time-changing mean fitness $vt$.
Defining $X = x - vt$, we then have $\partial_t p(x-vt) = -v\partial_X p(X)$.
This can be rewritten as a wave equation
\begin{equation}
v \partial_X p(X,s) = (X+s)p(X,s) - (1+X+s)p(X,s)^2.
\end{equation}
This is a Bernoulli equation and is readily solved, but the only appropriate boundary condition yields $p = 0$ everywhere.
This is intuitive: the inexorable advance of the mean fitness implies that all lineages are ultimately doomed to extinction \emph{unless} they can increase their own fitness either via recombination or mutation.
We thus need to amend the model to include a mutation term.
Let $U$ be the mutation rate of the genetic background and $\epsilon$ be the fitness effect of a mutation, distributed as $\rho(\epsilon)$.
Then we have
\begin{equation}
v \partial_X p(X,s) = (X+s)p(X,s) - (1+X+s)p(X,s)^2 + U\int_0^\infty \rho(\epsilon) \left[ p(X+\epsilon,s)-p(X,s)\right] d\epsilon,
\end{equation}
where $\epsilon$ is now the fitness effect of a mutation in the background distribution, $\rho(\epsilon)$ is the distribution of mutational effects, and $U$ is the mutation rate of the background.
Assuming that $s$ is small compared to $\sigma$, a matched asymptotics analysis yields
\begin{equation}
p(X,s) =
\begin{cases}
0 &\mathrm{~if~} X+s < 0, \\
X_c e^{\frac{(X+s)^2-X_c^2}{2\sigma^2}} &\mathrm{~if~} 0 < X+s < X_c, \\
X+s &\mathrm{~if~} X+s > X_c.
\end{cases}
\end{equation}
(If $s$ is large compared to $\sigma$, then the standard Haldane result, $p(s) \sim s$, obtains.)
Effectively, this means the fixation probability is \emph{zero} if the double mutant lineage's fitness is less than the average for the population: if the fitness is less than some critical value $X_c$, then the lineage must generate a background mutant in order to fix, and if it is above $X_c$, the fixation probability is determined by the lineage's ability to survive drift and establish.
The value of $X_c$ is determined by the shape of $\rho(\epsilon)$.
Suppose $\rho$ is distributed exponentially: $\rho(\epsilon) = \frac{1}{\bar{\epsilon}} e^{-\epsilon/\bar{\epsilon}}$.
Then $X_c$ is determined by solving
\begin{equation}
2 = \frac{\sqrt{2\pi}\sigma U}{\bar{\epsilon}} \left[ \frac{1}{X_c - \frac{\sigma^2}{\bar{\epsilon}}} + \frac{\bar{\epsilon}}{\sigma^2} \left( 1+ \frac{\bar{\epsilon}}{X_c} \right) \right] e^{\frac{(X_c - \sigma^2/\bar{\epsilon})^2}{2\sigma^2}}. 
\end{equation}

We now consider the behavior of single mutant bubbles.
Since the tunneling probability for a lineage with background fitness $x$ is given by $\phi = \left< 1 - e^{-\mu p_{\mathrm fix} W} \right>$, we do not need to compute the full distribution of $W$ but rather only its Laplace transform: $\mathcal{L}\left[ p(w) \right] = \int_0^\infty e^{-zw} p(w) dw$, with $z = \mu p_{\mathrm {fix}}$.
We proceed to derive an equation for the time evolution of the Laplace transform.

@@

To test our model, we perform forward simulations using a customized version of \texttt{FFPopSim} \citep{zanini_2012}, a discrete time forward simulation package.
The simulation code is available from the authors upon request or via \texttt{GitHub}.
We initialize a wild type population of $N$ individuals consisting of haploid genomes of length $L = 200$.
For the genetic background, we make use of an \emph{ersatz} ``infinite sites" model.
Any time a locus becomes monomorphic, a mutation is injected into a random individual in the population.
Hence, the population experiences a constant influx of beneficial mutations at the background loci.
Background fitness effects are drawn from an exponential distribution, and the background fitness variance is set to a constant value $\sigma^2$ by manually adjusting the selection coefficients every generation: they are multipled by the current variance and divided by $\sigma^2$.
In this way, $U$ and $\bar{\epsilon}$ are not parameters of our simulation model but must be directly measured from simulations: they are constrained by $L$ and $\sigma$.
We estimate them by manually counting the number of injected mutations every generation and averaging over the fitness effect at each locus.

@@

We partitioned our simulations as follows.
First, we considered the process of sequential fixation, in which the valley crossing time is dominated by the time until a deleterious mutant that is destined to fix arises: once it does, it does so in order $T_{\mathrm{MRCA}}$ generations, which is of order $N$ for low $N\sigma$ and scales weakly with $N$ for higher $N\sigma$ \citep{neher_hallatschek_2013}.
In the drift limit, this obtains when $N \ll \mathrm{min}(1/\delta, 1/\sqrt{\mu s})$, so that a successful mutant is unlikely to arise until the intermediate has fixed, but the intermediate's fitness disadvantage is too weak for drift to inhibit it.
Next, we considered the stochastic tunneling regime.
Tunneling becomes relevant when $N$ is too large for sequential fixation but also $\ll 1/\mu$, meaning that bubbles are unlikely to overlap and, if they do overlap, are unlikely to interfere with each other.
By increasing $N$ far above the $1/\mu$ limit, multiple deleterious mutants are likely to occur every generation, meaning that double mutants are produced at a high rate (semi-deterministic tunneling).

@@

\section*{Discussion}

We have used simulation methods to examine the rate at which rapidly adapting populations cross fitness valleys.
Though the overall time required is generally higher than in slowly adapting (drifting) populations, there are two important caveats.
First, the effect of the deleterious intermediate is much smaller.
Second, the time to the evolution of even \emph{simple} adaptations (i.e., ones requring substitution only at a single locus) is likely to be much longer in rapidly adapting populations.
This means that the presence of a fitness valley is not as steep of an impediment to the evolution of a complex adaptation for a drafting population as it is for a drifting one.
One may compare our results to those of \citet{ochs_2015}, which found that valley crossing in drifting populations is slowed by intermediate population sizes.

We have focused exclusively on asexual populations, but extensions to sexual populations are possible.
The relevant parameter is likely to be the product of $N$ and the proportion of the fitness standard deviation segregating in a small, effectively asexual block, $\sigma_b$ \citep{neher_kessinger_2013}.
If both $N\sigma$ and $N\sigma_b$ are large, and the focal loci segregate in the same block, then the analysis presented here should determine the crossing time.
If, on the other hand, $N\sigma$ is large but $N\sigma_b$ is small, or the distance between the focal loci is much larger than the block length $\xi_b$, then recombination between the focal loci is likely to be effectively ``free".
In that case, the communal recombination analysis of \citet{neher_shraiman_2011} and \citet{neher_shraiman_2010} is likely to be important.
The rate at which complex adaptations arise and fix will depend sensitively not on the distribution of background mutation effects but on the dynamics of recombination.
